Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-28 manuscript/supplemental_material.Rnw
move some tables and figures to supplemental material

@@

\section{Sequence Read Archive}

@@

@@

<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 5\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts}\n\n"))
    this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 5\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts.")
    print(xtable(go.results[grepl("Top 5% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                            list(GO.ID,Term,FDR,Ontology)
                            ][order(FDR)],
                 caption=this.caption,
                 label=paste0("tab:go_terms_enriched_top_five_",gsub(" ","_",this.species)),
                 display=c("s","s","s","e","s")
                 ),
          size="\\small",
          booktabs=TRUE, 
          tabular.environment="longtable",
          caption.placement="top",
          floating=FALSE)
}
@

@@

\section{Gene Expression}
<<top.expressed.genes,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    this.caption <- paste0("\\textit{",
                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
    cat(paste0("\\subsection{",this.caption,"}\n\n"))
    print(xtable(genes.expression.table(species.name=this.species),
                 caption=this.caption,
                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                 ),
          booktabs=TRUE,
          tabular.environment="longtable",
          floating=FALSE,
          caption.placement="top",
          sanitize.text.function=genes.table.sanitize,
          size="\\tiny"
          )
}
@

>>>>

\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{N. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}

\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
  expression of genes related to #1. The gene tree was obtained from
  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
  not included in this study or having expression by \ac{FPKM} less
  than 1 were excluded from the plotted tree. Heatmap cells in grey
  are from species where there was no expression of the relevant
  transcript (or a species where that transcript does not apply).}

@@

\section{Species}

\subsection{Sequence Source}

@@

\subsection{Species Alignment}

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 1\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts}\n\n"))
    this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                           " the top 1\\% of \\textit{",
                           capfirst(this.species),"} expressed transcripts.")
    print(xtable(go.results[grepl("Top 1% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                            list(GO.ID,Term,FDR,Ontology)
                            ][order(FDR)],
                 caption=this.caption,
                 label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
                 display=c("s","s","s","e","s")
                 ),
          size="\\small",
          booktabs=TRUE, 
          tabular.environment="longtable",
          caption.placement="top",
          floating=FALSE)
}
@

@@

\section{Gene Expression}

\subsection{Gene Evolution Trees}

\begin{figure}
<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of \textit{HBE1} and related genes}
\label{fig:hbe1_evolution_tree}
\end{figure}

\subsection{Clade specific Expression}

\begin{table}
  \footnotesize
<<lau.lineage.specific,echo=FALSE,results="asis">>=
lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
setkey(lau.all.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed in the Laurasiatheria
  superorder as compared to all other analyzed species.}
\label{tab:laurasiatheria_lineage_specific}
\end{table}

\begin{table}
  \footnotesize
<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
setkey(eua.lau.lin.spec,"Human Symbol")
@ 
\caption{Genes whose expression changed between the
  Euasiatherian and Laurasiatherian superorder.}
\label{tab:laurasiatheria_eurasiatherian_lineage_specific}
\end{table}

\subsection{Species Expression}

<<top.expressed.genes,echo=FALSE,results="asis">>=
for (this.species in names(species.names)) {
    this.caption <- paste0("\\textit{",
                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
    cat(paste0("\\subsection{",this.caption,"}\n\n"))
    print(xtable(genes.expression.table(species.name=this.species),
                 caption=this.caption,
                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                 ),
          booktabs=TRUE,
          tabular.environment="longtable",
          floating=FALSE,
          caption.placement="top",
          sanitize.text.function=genes.table.sanitize,
          size="\\tiny"
          )
}
@

\section{Core Placenta Transcriptome}

\begin{figure}
<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
calculate.distribution.shape <- 
    function(min.vector) {
        min.vector <- sort(min.vector,decreasing=TRUE)
        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
        return(temp)
    }

cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]

cpt.distribution.shape <- 
    lapply(c("min_expression",
             "second_min_expression",
             "third_min_expression",
             "fourth_min_expression",
             "fifth_min_expression"),
           function(x){
               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
           })
cpt.distribution.shape <- 
    rbindlist(cpt.distribution.shape)

cpt.distribution.shape <- 
    cpt.distribution.shape[,`Lowest Expression`:=
                               gsub("^min_expression$","First",
                                    `Lowest Expression`)]
cpt.distribution.shape <- 
    cpt.distribution.shape[,`Lowest Expression`:=
                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
                                           `Lowest Expression`,perl=TRUE),
                                      levels=c("First","Second","Third","Fourth","Fifth"))
                           ]
setnames(cpt.distribution.shape,
         "FPKM",
         "Expression (FPKM)")

print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
      + geom_line(lwd=1)
      + xlab(expression(FPKM))
      + ylab("Genes with N-th Lowest Expression")
      + scale_y_log10()
      + scale_x_log10()
      + geom_vline(xintercept=10)
      + geom_segment(mapping=aes(y=V1,
                         yend=V1,
                         xend=10,
                         x=100,
                         color=`Lowest Expression`),
                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                                              max(num_genes),`Lowest Expression`],
                     arrow=arrow(length=unit(0.5,"cm"))
                   )
      + geom_text(mapping=aes(x=100,
                      y=V1,
                      label=paste0(V1," genes"),override=TRUE),
                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
                      max(num_genes),`Lowest Expression`],
                  show_guide=FALSE,
                  hjust=0
                  )
      + guides(color=guide_legend(title="Lowest\nExpression"))
      )
@ 
\caption{Distribution shape of placenta core genes. The vertical axis shows
  the number of orthologous gene sets included in the core placenta
  transcriptome if the least expression cutoff (or second-least, etc.)
  was set to the horizontal axis. For example, if the cutoff is set to 10
  (vertical black line), 37 genes would be included if the FPKM of all
  species must be at least 10, 134 genes would be included if all but
  one species must have FPKM of at least 10. 655 genes would be
  included if all but five species must have FPKM of at least 10.}
\label{fig:core_placenta_transcriptome_shape}
\end{figure}
